On the propriety of restricted Boltzmann machines

### 2016-10-19 writing/draft.Rmd
first part of edits from Nordman

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created. RBMs are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degenerate models as well as the difficulties that arise in fitting such highly flexible models in the binary case.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse. In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data is generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. RBMs often fail to generate data that resemble realistic data and thus an unsatisfactory conceptualization of the data generation process. Additionally, we find that RBMs easily exhibit a kind of instability in the parameter space. In practice, this is seen when a single pixel change in an image results in a wildly different classification. Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep]. We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable. But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs. In section \ref{bayesian-model-fitting} we compare three Bayesian modelling techniques, each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As could be expected, with computational complexity comes an increase in fitting accuracy, but at the cost of feasibility. Additionally, because these models are highly overparametrized, any fully principled fitting method will reproduce the empirical distribution of most realistic training sets. There is no "smoothing" achieved with a RBM fit using a rigorous method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability in section \ref{appendix}.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common. In this paper, we consider the binary case of an RBM only. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no connections within a layer. An example of this structure is in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

@@

A common use for RBMs is to create features for use in classification. For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$.

### Joint distribution
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, two binary $H + V$-dimensional spaces. We posit a parametric form for probabilities corresponding to a potential vector of states of each node taking the value of $1$.

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} 
\end{align}

Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function. The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion. A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of the RBM. Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the main and hidden layers.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) makes three kinds of potential model impropriety of concern, *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationship to one another.

>>>>

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degeneracy, instability and/or uninterpretability in such models in the binary case. We also describe the difficulties that arise in likelihood-based fitting such highly flexible models, especially as Gibbs sampling methods are often advocated for the RBM model structure.
---

@@

The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs). (**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information]. These models have reportedly been used with success in classification of images. (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse (**TODO: citation**). In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology. We find the RBM model class to be deficient in two fundamental ways.

First, the models are unsatisfactory as a conceptualization of how data are generated. According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical]. Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination]. RBMs can often generate data lacking realistic variability and thereby fail to provide a satisfactory conceptualization of the data generation process. In addition to such degeneracy, we find that RBMs easily exhibit types of instability in the parameter space. In practice, this may seen when a single pixel change in an image results in a wildly different classification. Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep]. We investigate these phenomena in RBMs in Section \ref{explorations} through simulations of small, manageable examples.

Secondly, the fitting of these models is problematic. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable. But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs. The literature often suggests MCMC approximations to fit these models (exploiting conditional structure in the hidden and visible variables), but little is said about monitoring convergence [hinton2010practical, @hinton2006fast]. Related to this, the MCMC algorithms involve updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods. In Section \ref{bayesian-model-fitting}, we compare three Bayesian modelling techniques (which can be computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models. As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of feasibility. Additionally, because these models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the empirical distribution of most realistic training sets. There is no "smoothing" achieved with a RBM fit using a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.

This paper is structured as follows, Section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{bayesian-model-fitting} and conclude with a discussion in Section \ref{discussion}. Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common. In this paper, we consider the binary case of an RBM for concreteness. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no connections within a layer. An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

@@

A common use for RBMs is to create features for use in classification. For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$ [@hinton2006fast].

### Joint distribution
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, as the potential encodings for a binary $H + V$-dimensional space, $\mathcal{X}$. We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{X} 
\end{align}

where $\boldsybol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathcc{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function. Note that the RBM model is parameterized by $\boldsymbol \theta$containing two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion. A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of conditioal independence of the RBM (i.e. visibles given hiddens or vice versa). Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the visible and hidden layers.

@@

## Degeneracy, instability, and uninterpretability... Oh my!
The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates three kinds of potential model impropriety of concern: *degeneracy, instability, and uninterpretability*. In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationships to one another.
